Repeatability of Pre-Flashover Fire Patterns on Gypsum Wall Board

### 2015-03-15 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Kevin and Craig ignore  this version working several sections at once - don't ask.

A free standing, non-combustible, calcium silicate wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

A free standing, non-combustible, steel frame with gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

\chapter{Instrumented Calibration (non-combustible) Wall Experiments}
Now that the characteristics of the fires have been measured, the next step is to measure the thermal impact that the fires have on a target.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to eliminate any contribution from the wall to the fire.  The fires were initially placed at predetermined, fixed positions away from the wall.  In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610 m), one burner width (0.305 m), one half burner width (0.150 m), and the touching the wall.

\section{Experimental Arrangement}
A free standing steel frame wall with a 12.7 mm thick, 2.44 m high by 1.22 m wide, calcium silicate face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2 m to 1.2 m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20 cm intervals, up to a height of 2.2 m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall of the wall.  The bead of the thermocouple was flush with the surface on the calcium silcate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10 mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory’s 3~m by 3~m oxygen depletion calorimeter.

@@

\chapter{Instrumented Gypsum Board Wall Experiments}
Purpose
\section{Experimental Arrangement}
A free standing, non-combustible, steel frame with gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

@@

@@

@@

\section{Wall Construction Type Experiments}
\subsection{Experimental Arrangement}
\subsection{Results}

>>>>

A free standing, steel frame with non-combustible, calcium silicate faced wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

\chapter{Instrumented Calibration (non-combustible) Wall Experiments}

Now that the characteristics of the fires have been measured, the next step is to measure the thermal impact that the fires have on a target.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to eliminate any contribution from the wall to the fire.  The fires were initially placed at predetermined, fixed positions away from the wall.  In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610 m), one burner width (0.305 m), one half burner width (0.150 m), and the touching the wall.

A free standing steel frame wall with a 12.7 mm thick, 2.44 m high by 1.22 m wide, calcium silicate face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2 m to 1.2 m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20 cm intervals, up to a height of 2.2 m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall of the wall.  The bead of the thermocouple was flush with the surface on the calcium silcate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10 mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory’s 3~m by 3~m oxygen depletion calorimeter.

@@

\chapter{Instrumented Gypsum Board Wall Experiments}
Purpose
\section{Experimental Arrangement}
A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

The gypsum wallboard used in all of these experiments was purchased at one time from one vendor in an effort to have the most similar panels for used between the experiments.  The physical features of the gypsum wallboard panels are given in the table below.   After receipt of the gypsum wallboard panels, the panels were laid out on the floor or the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer.  Once the primer dried a top coat of white latex paint was applied.  After the panels dried paint thickness measurements were made with a ultra sonic XXXX.  Some of the panels were then cut-up into samples     think about this   perhaps this goes into the FP Comp

@@

The cone calorimeter is probably the most versatile oxygen consumption method for measuring heat release rate. It was developed at NIST in the 1980s43 and is presently the most commonly used bench-scale rate of heat release apparatus.44 The cone calorimeter has been adopted as ASTM E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter.19 It has also been adopted as an international standard, ISO 5660-1.45
The cone calorimeter consists of a heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5 kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25 mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100 kW/m2. Figure  is a schematic of the cone calorimeter.

For piloted ignition tests, an electric spark ignitor is positioned at the top of vertical samples and over the center of horizontal samples. Samples are typically 100 mm by 100 mm, and they can be wrapped with aluminum foil to minimize edge effects. Combustion products and dilution air are extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03 m3/sec. The volumetric flow rate is kept constant during testing. The sample is mounted on a load cell to determine mass loss rate during a test, and smoke obscuration is measured using a laser light source. The gas flow rate in the exhaust duct is calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, and other gases are measured using appropriate instruments. Heat release rate is calculated from the gas concentration and mass flow measurements.

For this study samples were cut from painted gypsum wallboard panels that were prepared with the panels used in the compartment experiments. Key measurements made with the cone calorimeter, in addition to heat release rate per unit area of the samples, included using a series of cone calorimeter experiments at different heat fluxes to determine a ignition time and surface temperature pairs.

@@

Normally the radiant panel test is used as documented in ASTM~E162 and provides a measure of downward, opposed-flow flame spread..63  The apparatus measures the surface flammability of materials using a gas-fired radiant panel that is 305~mm × 457~mm. However for the purposes of this study a modified version of the test apparatus was used.  The radiant panel and and the radiant panel controls remained in place as documented in ASTM~E162. However the sample mounting system was changed. A rail and trolley system was developed to expose fire fighter protective equipment to various heat fluxes by being able to position the test specimen to a pre-determined position that correlated with a given heat flux.  This trolley was used to hold a sample of painted gypsum wallboard were exposed to the radiant .   Figure

The exposed surface of the samples was parallel to the radiant panel.  The radiant panel was calibrated and mapped with a heat flux gauge. With the flux gauge data, positions away from the radiant panel were identified to enable the samples to be exposed to a range of heat fluxes.  After a timed exposure the samples were examined and amount of thermal damage to the painted paper surface of the gypsum board was   A relative index of flame spread is calculated based on the distance the sample burns. This test method is referenced for building products as ASTM E162, Standard Test Method for Surface Flammability of Materials using a Radiant Energy Source64 and for cellular plastics as ASTM D3675, Standard Test Method for Surface Burning Characteristics of Building Materials.65 Figure 2.3.11 illustrates the radiant panel test apparatus.

@@

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90 mm by 38 mm.  A partial ceiling with a width of at least 1.2 m, measured from the back wall of the compartment was installed across the width of the compartment, this allowed for the plume to impact the ceiling and form a ceiling jet along the back wall, before the fire gases flowed up and out of the compartment and into the 6~m by 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7 mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2 m wide and 2.4 m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway. These experiments were conducted without instrumentation in the walls for several reasons; to compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation, to provide more realistic conditions, and to enable the conduct of additional replicate experiments given the limited laboratory availability.  Based on the outcome of the instrumented calibration target wall results, burner positions were chosen. After each burn, the fire patterns were measured, photographed and analyzed for repeatability. The fire damaged wallboard was subsequently removed and replaced with new sections of painted gypsum wallboard. The number of experiments for each case was dependent on the repeatability and similarity of the pattern development. Several types of fire pattern experiments were conducted in the compartments, as detailed in the sections to follow, to examine the impact of the insulation behind the gypsum wallboard panel on the resulting fire pattern and other series of experiments were conducted to compare the fire patterns based on the burner centered on the rear wall or in or near a corner on the rear wall of the compartment.

\section{Wall Construction Type Experiments}

 Impact of Wall Construction 
 Additional experiments were conducted with the natural gas and gasoline fires to examine the impact of the method of construction of the walls on the development of the fire patterns.  Three cases were examined: 1) Open back, 12.7 mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated,12.7 mm thick, regular core, painted gypsum board on the front and back of a wood frame, with fiberglas batt insulation in the voids of the wood frame representative of an exterior wall.

The fuel or burner was centered on the middle panel of the rear wall.  As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5 cm above the floor of the compartment.  Each of the painted gypsum board panels that were used for a fire pattern had a 5 cm grid drawn on it to assist with the determination of the geometry of the fire pattern.

\subsection{Experimental Arrangement}
\subsection{Results}
